Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-02-21 writeups/Su_IGW_break.tex
everything is done except for movie plots now

% cp ../sims/agg.png plots; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/f_amps.png plots/lin_amps.png; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/fluxes.png plots/lin_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps2.png plots/nl_f_amps2.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps.png plots/nl_f_amps.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/fluxes.png plots/nl_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_refl.png plots/nl_f_refl.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_ri.png plots/nl_f_ri.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/front.png plots/nl_front.png;

@@

\subsection{Wave Generation}

@@

The geometry of our simulation domain is fixed by one further parameter, $z_0$,
the forcing location. We choose $L_z = 12.5H$, and the lower and upper damping
zones are located at $z < 0.3H$ and $z > 9.5H$ respectively ($z \in [-1,
11.5]$). The forcing (see Eq.~\eqref{eq:vol_drive}) is at $z_0 = 2H$ with width
$\sigma = 0.078H$, sufficiently far from the lower damping zone and permitting
sufficient room for the upward propagating wave to grow as $\propto e^{z/2H}$.
Finally, we want similar grid spacing in the $x$ and $z$ directions (i.e.\ $L_x
/ N_x \sim L_z /N_z$), guided by the intuition that turbulence generated by wave
breaking is approximately isotropic, so we use $L_x = 4H$ and $N_z / N_x = 4$.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 1.5H$ and $z_T = 10.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. We choose the transition width $\Delta z = 0.25H$ and damping
time $\tau = 1 / (15N)$. This prescription is similar to \citet{lecoanet_damp}
and has the advantage of being smooth, important for spectral methods. Further
details of our implementation of the fluid equations in Dedalus are described in
Appendix~\ref{se:strat_impl}.

@@

\begin{figure*}
    \includegraphics[width=0.45\textwidth]{plots/yubo_000054.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000086.png}

    \includegraphics[width=0.45\textwidth]{plots/yubo_000153.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000451.png}
    \caption{Snapshots of $u_x$ (in units of $HN$) and $\Upsilon \equiv
    \ln\p{\rho / \bar{\rho}}$ in the fiducial simulation illustrating distinct
    phases of the evolution of the flow. Note that $\overline{U}_c \approx
    0.16HN$ (see Eq.~\eqref{eq:u_crit}) for the parameters used. The four times
    depicted are $t = 413.4/N$ (top left), $t = 658.5/N$ (top right), $t =
    1171.4/N$ (bottom left), and $t = 3437.8/N$ (bottom right). The four
    snapshots illustrate (i) the initial transient wave breaking phase, (ii)
    formation of a distinct critical layer, (iii) steepening of the critical
    layer, and (iv) downward advance of the critical
    layer.}\label{fig:snapshots}
\end{figure*}

@@

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
We use the background buoyancy frequency to compute $\mathrm{Ri}$, as
fluctuations do not change $N^2$ significantly ($\sim 1\%$). To understand the
variation in $\mathrm{Ri}$ over $x$, we also compute $\min\limits_x
\mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown in
Fig.~\ref{fig:nl_f_ri}. The Richardson number at the critical layer quickly
decreases to a typical value between $0.25$ and $0.5$, characteristic of the
onset of the KHI\@.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number (Eq.~\eqref{eq:ri_med_def}) of the flow at
    the critical layer over time (in units of $N^{-1}$) in our fiducial
    simulation. The red and solid green lines denote respectively the minimum
    and median of $\mathrm{Ri}_x(x, t)$. These numbers measure the mean and
    spread in width of the critical layer over $x$. Note that $\mathrm{Ri} \sim
    \frac{1}{4}$ corresponds to the KHI, so this plot suggests the shear at the
    critical layer does not steepen past the onset of the
    KHI.}\label{fig:nl_f_ri}
\end{figure}

The steadiness of $\mathrm{Ri}$ at this characteristic value suggests that the
critical layer width is self-consistently determined by the ``turbulence'' in
our simulations. The width does not change significantly with resolution in our
resolved simulations (see Fig.~\ref{fig:agg}). Thus, in stars, $\mathrm{Ri} \sim
0.5$ can be used to calculate the critical layer width, where $N^2$ and
$\overline{U}_c$ (corresponding to the tidal frequency) are known.

@@

It is interesting to compare our work with that of \citet{barker_ogilvie}, who
studied inward-propagating IGWs in solar-type stars and their nonlinear breaking
due to geometric focusing. In their numerical simulations in a 2D polar
geometry, they found no evidence for reflected waves, contrary to our result.
% It is plausible that the discrepancy arises due to the differing causes of wave
% breaking (geometric focusing vs.\ density rarefaction). Note that their
% nonlinearity parameter $\lambda \sim 1/\mathrm{Re}$ where $\mathrm{Re}$ is the
% Reynolds number (Eq.~\eqref{eq:re_def}). For $\lambda \sim 10^{-1}$ as explored
% in \citet{barker_ogilvie}, our results also show no reflected wave.
Note, however, that their simulations were run with substantially higher
viscosity, i.e.\ lower resolution (their $\lambda \sim 1/\mathrm{Re}$, so they
studied $\mathrm{Re} \sim 10$), than the simulations presented here. We also
find at low Reynolds number that there is negligible wave reflection. It is
possible that reflection and flux redistribution are also important processes
for geometrically breaking waves, but was absent from the low-Reynolds number
simulations of \citet{barker_ogilvie}.

>>>>

% cp ../sims/agg.png plots; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/f_amps.png plots/lin_amps.png; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/fluxes.png plots/lin_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps2.png plots/nl_f_amps2.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps.png plots/nl_f_amps.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/fluxes.png plots/nl_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_refl.png plots/nl_f_refl.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_ri.png plots/nl_f_ri.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/front.png plots/nl_front.png;
% cp ../sims/2d_4_fourier/frames2/*labeled.png plots

@@

\subsection{Wave Generation}\label{ss:forcing}

@@

The geometry of our simulation domain is fixed by one further parameter, $z_0$,
the forcing location. We choose $L_z = 12.5H$ ($z$ runs from $-H$ to $11.5H$),
and the lower and upper damping zones are active for $z < 0.3H$ and $z > 9.5H$
respectively. The forcing (see Eq.~\eqref{eq:vol_drive}) is at $z_0 = 2H$ with
width $\sigma = 0.078H$, sufficiently far from the lower damping zone and
permitting sufficient room for the upward propagating wave to grow as $\propto
e^{z/2H}$. Finally, we want similar grid spacing in the $x$ and $z$ directions
(i.e.\ $L_x / N_x \sim L_z /N_z$), guided by the intuition that turbulence
generated by wave breaking is approximately isotropic, so we use $L_x = 4H$ and
$N_z / N_x = 4$.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. We choose the transition width $\Delta z = 0.25H$ and damping
time $\tau = 1 / (15N)$. This prescription is similar to \citet{lecoanet_damp}
and has the advantage of being smooth, important for spectral methods. Further
details of our implementation of the fluid equations in Dedalus are described in
Appendix~\ref{se:strat_impl}.

@@

\begin{figure*}
    \includegraphics[width=0.45\textwidth]{plots/yubo_000040_labeled.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000089_labeled.png}

    \includegraphics[width=0.45\textwidth]{plots/yubo_000134_labeled.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000451_labeled.png}
    \caption{Snapshots of $u_x$ and $\Upsilon$ in the fiducial simulation
    illustrating distinct phases of the evolution of the flow. Damping layers
    are shaded in light grey (see Section~\ref{ss:damping}), while the forcing
    zone (see Section~\ref{ss:forcing}) is shaded in light green (boundaries are
    at $z_0 \pm 3\sigma$). The four panels illustrate (i) the initial transient
    wave breaking phase, (ii) formation of a distinct critical layer, (iii)
    steepening of the critical layer, and (iv) downward advance of the critical
    layer.}\label{fig:snapshots}
\end{figure*}

@@

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
We use the background buoyancy frequency to compute $\mathrm{Ri}$, as
fluctuations do not change $N^2$ significantly ($\sim 1\%$). To understand the
variation in $\mathrm{Ri}$ over $x$, we also compute $\min\limits_x
\mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown in
Fig.~\ref{fig:nl_f_ri}. Absorbtion of incident IGWs quickly decreases the
Richardson number to between $0.25$ and $0.5$, characteristic of the onset of
the KHI\@.

This result suggests that the critical layer width is regulated by the
competition between steepening induced by IGW breaking and broadening due to
shear instability. This width does not vary significantly with resolution in our
resolved simulations (see Fig.~\ref{fig:agg}). $\mathrm{Ri} \sim 0.5$ can be
used to calculate the critical layer width in stars, where $N^2$ and
$\overline{U}_c$ (corresponding to the tidal frequency) are known.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number (Eq.~\eqref{eq:ri_med_def}) of the flow at
    the critical layer over time (in units of $N^{-1}$) in our fiducial
    simulation. The red and solid green lines denote respectively the minimum
    and median of $\mathrm{Ri}_x(x, t)$. These numbers measure the mean and
    spread in width of the critical layer over $x$. Note that $\mathrm{Ri} \sim
    \frac{1}{4}$ corresponds to the KHI, so this plot suggests the shear at the
    critical layer does not steepen past the onset of the
    KHI.}\label{fig:nl_f_ri}
\end{figure}

@@

It is interesting to compare our work with that of \citet{barker_ogilvie}, who
studied inward-propagating IGWs in solar-type stars and their nonlinear breaking
due to geometric focusing. In their numerical simulations in a 2D polar
geometry, they found no evidence for reflected waves, contrary to our result.
Note that their simulations were run with substantially higher viscosity, or
lower resolution, than explored here, and their effective Reynolds number (equal
to $1/\lambda$ in their notation) is of order $10$. We also find at low Reynolds
numbers that there is negligible wave reflection.
